Synthesizing Quotient Lenses

### 2017-10-05 papers/QuotientLensSynth/main.tex
Progress with Solomon on the QRE paper

\section{Introduction}
Lense programming languages such as Boomerang enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
program. For instance, lenses enable a programmer to write a program which
enables a doctor to acquire a summary of a patient's data, and also allows the
doctor to reflect updates made to the patient's record back into the main
database.

Lenses therefore enable the  programmer to kill two birds in one
stone by allowing the programmer to express two different data transformations
in one program. Also, when lenses are appropriately constrained, they offer
guarantees about how the two transformations interact with each other. Lenses
have therefore been used effectively as a solution to the \textit{view update
problem}, a classical problem in the database community in which the
programmer wishes to write a program which provides a \textit{view} of data
derived from a \textit{source} to a client, as well as a program which enables
the client to reflect changes made in the view back into the source without
corrupting the source.

However, the constraints that are imposed on lenses are often too strict.
For instance, the programmer may wish to ignore updates to the data that only
change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses} as a
solution to this problem. Quotient lenses enable a programmer to write lenses
which transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

@@

For example, suppose that the programmer wishes to convert citation records
stored in \textsc{Bib}\TeX{} format to a representation of the same data in the
EndNote format. The programmer wishes to ``quotient'' away unnecessary
whitespace characters in between the various fields in each record, while also
ignoring the order in which these fields occur in the recore. For instance, the
quotient lens which the programmer wishes to derive should convert this
\textsc{Bib}\TeX{} record:

@@

to this EndNote record

@@

\noindent while allowing for this equivalent representation of the \textsc{Bib}\TeX{}
entry, which rearranges the fields and uses different whitespace conventions:

\begin{verbatim}
@Book{conway, 
    Publisher = {Printed in GB by William Clowes & Sons Ltd},
        Title = {Regular Algebra and Finite Machines},

         Year = {1971},
       Author = {Conway, J. H.},
}
\end{verbatim}

@@

The second challenge that we address in this paper is the synthesis of
\textit{bijective} quotient lenses. These are quotient lenses which match the
equivalance classes of the source data to the equivalence classes of the view
data bijectively as in the \textsc{Bib}\TeX{} to EndNote example above. 
Building on our prior work~\cite{popl18} where we showed how to synthesize
(ordinary?) bijective lenses, we show how to synthesize bijective quotient
lenses given a synthesis task described by a pair of QREs and a set of
examples. To our knowledge this is the first time that this has been done.

Our plan for this paper is as follows:
\begin{enumerate}
  \item
  We introduce the language of Quotient Regular Expressions (QREs) which enables
  us to express a set of equivalence relations on regular languages using a
  compact, easy-to-use syntax.
  \item
  We describe a class $\mathcal{Q}$ of quotient lenses whose types are given by
  QREs. We derive a normal form for quotient lenses in $\mathcal{Q}$, and
  demonstrate that elements of $\mathcal{Q}$ that are in this normal form are
  closed under left quotienting, right quotienting, composition and regular
  operators.
  \item
  Using the normal form for quotient lenses in $\mathcal{Q}$, we reduce the
  synthesis of quotient lenses in $\mathcal{Q}$ to the synthesis of bijective
  lenses between regular languages. 
  \item
  We describe an implementation of QREs, the language $\mathcal{Q}$ of quotient
  lenses, and an algorithm for synthesizing quotient lenses in $\mathcal{Q}$ in
  the Boomerang language.
  \item
  We assess the usabibility of the QRE synthesis via a case study in which we
  selected several data formats used by data.gov and synthesized quotient lenses
  between them.
\end{enumerate}

\section{Quotient Regular Expressions}
This section introduces Quotient Regular Expressions or QREs. QREs are regular
expressions augmented with syntax that lets them simultaneously express an
equivalence relation on the language described by the regular expression.

\sam{More explanation here?}

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
  \subsection{Meaning and Example Usages for QREs}
  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. 
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions.}
  \label{fig:wk}
\end{figure}

@@

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended meaning:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $EqRel(q)$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$,
  \item $Canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $Canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $EqRel(q)$.
\end{theorem}
\section{Using QREs to Express Quotient Lenses}
Given regular expressions $R, S$ and equivalence relations $\sim_R, \sim_S$ are
defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$ respectively, a
\textit{quotient lens} $q :
R/{\sim_R}{\Longleftrightarrow} S/{\sim_S}$ from $R$ to $S$ is
a pair of functions $q.get:
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that
q.put \; (q.get \; r) &\sim_R r\\
q.get \; (q.put \; s) &\sim_S s
Additionally the components of $q$ must respect $\sim_R$ and $\sim_S$ i.e.
if $r \sim_R r'$ and $s \sim_S s'$ then $q.get \; r \; \sim_S q.get \; r'$
and $q.put \; s \sim_R q.put \; s' \; s'$.\\
Now let $c, c'$ be QREs. Since $K(c)$ and $K(c')$ form a complete set of
representatives for $W(c)$ modulo $EqRel(c)$ and $W(c')$ modulo $EqRel(c')$
respectively, then given a bijection $\ell : K(c) \Leftrightarrow K(c')$, we
may define a quotient lens $q : W(c)/EqRel(c) \Leftrightarrow
W(c')/EqRel(c')$ by 
q.get &= \ell \circ Canonizer(c)\\
q.put &= \ell^{-1} \circ Canonizer(c')
Define the class of \textit{bijective
lenses} to be the set of bijections between regular languages created from Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := R \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.\\
The denotation of a bijective lens $\ell$ is a function from 
$\llbracket e \rrbracket$ of a bijective lens $\ell$ is defined
inductively by
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}
\textcolor{red}{[More explanation needed esp. inference rules for bijective
lenses]}\\
Having described the class of bijective lenses, we now define the class
$\mathcal{Q}$ of quotient lenses, whose language is defined inductively by the
following grammar:
$$ q := id(\ell) \sep lquot(q, \ell) \sep rquot(\ell, q) \sep q
\cdot q_2 \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses 
\subsection{Semantics of $\mathcal{Q}$}
Every element in $\mathcal{Q}$ has a type $q :
c \Longleftrightarrow c'$ where $c$ and $c'$ are QREs. The denotation
$\llbracket q \rrbracket$ of an element $q : c \Longleftrightarrow c'$ is a
quotient lens $\llbracket q \rrbracket :
W(c)/{EqRel(c)} \Longleftrightarrow W(c')/{EqRel(c')}$.\\
\textcolor{red}{[Informal Meaning/Usages Needed]}\\We now give the
semantics of $\mathcal{Q}$, as well as inference rules that may be used to
derive elements of $\mathcal{Q}$:
\begin{enumerate}
  if $q = id(\ell)$, then
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$id(\ell): R \Leftrightarrow id(S)$}
\end{prooftree}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}

@@

\section{Synthesizing Quotient Lenses}
By \cref{normal form}, if there is a derivation $q : c \Leftrightarrow c'$ in
$\mathcal{Q}$ then there exists a bijective lens $\ell : K(c) \Leftrightarrow
K(c')$ such that
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ Canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
In other words, every quotient lens in $\mathcal{Q}$ is the same as an ordinary
bijective lens with canonizers at the ends.\\
Therefore, in order to synthesize a quotient lens $q: c \Leftrightarrow c'$
where $c, c'$ are QREs, it suffices to synthesize a bijective lens $\ell : K(c)
\Leftrightarrow K(c')$.\\
\textcolor{red}{[Citation Needed!]} has already shown that if there
exists a derivation $\ell : R \Leftrightarrow S$, then there exist
regular expressions $R', S'$ and a synthesizable \\
\textcolor{red}{[Explain and make explicit the meaning
of synthesizable]} \\
bijective lens $\ell' :

>>>>

\section{Introduction}

{\em Lens programming languages} such as
Boomerang~\cite{Boomerang}\bcp{and others!} enable programmers to kill two
birds in one stone by expressing both directions of a bidirectional
transformation with a single program.

\bcp{
  \begin{itemize}
  \item Small example of a boomerang program (maybe in our syntax)
  \item Discussion of lens laws.
  \end{itemize}
}

However, the constraints that are imposed by the lens laws are sometimes too
strict. 
For instance, the programmer may wish to ignore updates to the data that only
change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses} as a
solution to this problem. Quotient lenses enable a programmer to write lenses
which transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

@@

For example, suppose that the programmer wishes to convert citation records
stored in \textsc{Bib}\TeX{} format to a representation of the same data in the
EndNote format. The programmer wishes to ``quotient'' away unnecessary
whitespace characters in between the various fields in each record, while also
ignoring the order in which these fields occur in the recore. For instance, the
quotient lens which the programmer wishes to derive should convert this
\textsc{Bib}\TeX{} record

@@

\noindent to this EndNote record

@@

\noindent and vice versa.

\bcp{Moreover...}
allowing for this equivalent representation of the \textsc{Bib}\TeX{}
entry, which rearranges the fields and uses different whitespace conventions:

\begin{verbatim}
@Book{conway, 
    Publisher = {Printed in GB by 
                 William Clowes & Sons Ltd},
        Title = {Regular Algebra and Finite Machines},
         Year = {1971},
       Author = {Conway, J. H.},
}
\end{verbatim}
\bcp{Can we write (regular and QRE) lenses that handle this version of the
  example?  Write this in Boomerang and see how it looks!!}

@@

\bcp{Give a little example (perhaps for Bibtex and/or EndNote?)}

The second challenge that we address in this paper is the {\em synthesis} of
(bijective) quotient lenses. These are quotient lenses which match the
equivalance classes of the source data to the equivalence classes of the view
data bijectively as in the \textsc{Bib}\TeX{} to EndNote example above. 
Building on our prior work~\cite{popl18} where we showed how to synthesize
(ordinary?) bijective lenses, we show how to synthesize bijective quotient
lenses given a synthesis task described by a pair of QREs and a set of
examples.

Our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a broad class of
equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a class of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from types and examples to the previously studied problem of
synthesizing bijective lenses between regular languages
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.go} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions}
\label{QRE}

This section introduces Quotient Regular Expressions or QREs. QREs are regular
expressions augmented with syntax that lets them simultaneously express an
equivalence relation on the language described by the regular expression.

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function\bcp{Words in math mode need $\backslash${\tt
      mathit} around them, always.} $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.

  \subsection{Examples}

\bcp{Give some counterexamples too!  And can we take away the enumerate?}

  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. \bcp{Suggest using $\backslash${\tt mathtt} for 
      concrete program identifiers and ascii characters in examples.}
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{left-justify columns,
      and try putting the last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

@@

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations\bcp{too wide -- remove LH column and
      try to get the last clause into the same table}}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers\bcp{First column not needed (and, once it's gone,
      maybe the = signs could be vertically aligned?)}}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended meaning:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $EqRel(q)$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$,
  \item $Canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $Canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $EqRel(q)$.
  \end{enumerate}
\end{theorem}

\section{QRE Lenses}
\label{QRE-lenses}

Given regular expressions $R, S$ and equivalence relations $\sim_R, \sim_S$ are
defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$ respectively, a
\textit{quotient lens} $q :
R/{\sim_R}{\Longleftrightarrow} S/{\sim_S}$ from $R$ to $S$ is
a pair of functions $q.get:
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that
\begin{align*}
q.put \; (q.get \; r) &\sim_R r\\
q.get \; (q.put \; s) &\sim_S s
\end{align*}
Additionally the components of $q$ must respect $\sim_R$ and $\sim_S$ i.e.
if $r \sim_R r'$ and $s \sim_S s'$ then $q.get \; r \; \sim_S q.get \; r'$
and $q.put \; s \sim_R q.put \; s' \; s'$.\\
Now let $c, c'$ be QREs. Since $K(c)$ and $K(c')$ form a complete set of
representatives for $W(c)$ modulo $EqRel(c)$ and $W(c')$ modulo $EqRel(c')$
respectively, then given a bijection $\ell : K(c) \Leftrightarrow K(c')$, we
may define a quotient lens $q : W(c)/EqRel(c) \Leftrightarrow
W(c')/EqRel(c')$ by 
\begin{align*}
q.get &= \ell \circ Canonizer(c)\\
q.put &= \ell^{-1} \circ Canonizer(c')
\end{align*}
Define the class of \textit{bijective
lenses} to be the set of bijections between regular languages created from Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := R \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.\\
The denotation of a bijective lens $\ell$ is a function from 
$\llbracket e \rrbracket$ of a bijective lens $\ell$ is defined
inductively by
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}
\end{align*}
\textcolor{red}{[More explanation needed esp. inference rules for bijective
lenses]}\\
Having described the class of bijective lenses, we now define the class
$\mathcal{Q}$ of quotient lenses, whose language is defined inductively by the
following grammar:
$$ q := id(\ell) \sep lquot(q, \ell) \sep rquot(\ell, q) \sep q
\cdot q_2 \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses 
\subsection{Semantics of $\mathcal{Q}$}
Every element in $\mathcal{Q}$ has a type $q :
c \Longleftrightarrow c'$ where $c$ and $c'$ are QREs. The denotation
$\llbracket q \rrbracket$ of an element $q : c \Longleftrightarrow c'$ is a
quotient lens $\llbracket q \rrbracket :
W(c)/{EqRel(c)} \Longleftrightarrow W(c')/{EqRel(c')}$.\\
\textcolor{red}{[Informal Meaning/Usages Needed]}\\We now give the
semantics of $\mathcal{Q}$, as well as inference rules that may be used to
derive elements of $\mathcal{Q}$:\bcp{I think the numbering and the ``if $q$
  = ..." stuff is not needed.  And maybe we can save space by typesetting the
  get/put parts on the same line as the typing rule, at least in some cases.}
\begin{enumerate}
  \item
  if $q = id(\ell)$, then
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$id(\ell): R \Leftrightarrow id(S)$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}
  \end{align*}

@@

\section{Synthesizing QRE Lenses}
\label{synth}
